The Structure of Human Language Reflects Limitations of Working Memory

### 2019-04-28 paper-short.tex
revising

\documentclass[11pt,letterpaper]{article}

\usepackage{showlabels}
%\usepackage{fullpage}
\usepackage{pslatex}
%\usepackage{latexsym}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{url}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{amssymb}

@@

\frenchspacing
%\def\baselinestretch{0.975}

%\emnlpfinalcopy
%\def\emnlppaperid{496}

@@

@@

@@

%Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.

@@

%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\citep{berwick1984grammatical,hawkins1994performance};
however, it has been hard to test this idea without ad-hoc modeling assumptions.
%Alternative?: there has been debate about what shapes human language. role played by UG/innate constraints? environment? cognition?

We directly test this hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\citep{nivre-universal-2017}.

\mhahn{smooth out}We consider an individual listener's language comprehension process, processing a stream of words uttered by an interlocutor.
Experimental research has established that listeners form probabilistic expectations about the upcoming words.
Consider example...
Human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context.
Surprisal is known to be a strong linear predictor of processing effort as measured by per-word reading times.
For an idealized listener with unbounded memory resources, surprisal will be given by the negative log-probability of a word given its preceding context:
\begin{equation}\label{eq:surp}
	-\log p(x_i|x_{1...i-1})
\end{equation}
where $x_i$ is the $i$-th word in a given utterance (or discourse).

In reality, a listener will have bounded capacity to represent prior context.
A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
We denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
This listener will experience surprisal
\begin{equation}\label{eq:surp-listener}
	S :=	-\log p(x_t|L_t)
\end{equation}
upon hearing the next word $x_t$.
%The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
%For a listener who invests less bits into representing information $L_t$ about prior context, $L_t$ will be a more coarse-grained representation of past observations.
%Thus, the predictions made by such a listener will be on average less accurate, and such a listener will incur greater processing effort $S$.

Information theory, specifically the Data Processing Inequality, implies that the average surprisal $S$ will be higher for a listener who represents less information about past input.
Consequently, there is a tradeoff between memory and surprisal:
A listener can achieve lower levels of processing effort, at the expense of increased memory load.

Formally, we quantify memory as the number of bits about prior input that a listener represents.

@@

%TODO
Let $I_t$ be the Conditional Mutual Information of two words $X_0, X_t$ conditioned on the intervening words:
\begin{equation}
	\operatorname{I}_t := \operatorname{I}[X_t, X_0 | X_1, \dots, X_{t-1}] = \operatorname{H}[X_t|X_1, \dots, X_{t-1}] - \operatorname{H}[X_t|X_0, \dots, X_{t-1}] 
\end{equation}
This quantity measures how much predictive information is provided by the next word $X_t$ by the word $t$ steps in the past.
It is a statistical property of the language, and can be estimated from large-scale text data.
Our theoretical results describes how the tradeoff between memory and comprehension difficulty relates to $I_t$.

Consider a listener who invests $M$ bits of memory into representing past input.
We then consider the smallest $T$ such the area under the curve of $t I_t$, to the left of $T$, has size $M$.
Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
By tracing out all values $T >0$, one can obtain a bound on the tradeoff curve for any possible listener.

@@

%We prove that, for any integer $T > 0$, a listener encoding $\sum_{t=1}^T t I_t$ bits of memory will incur average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.

%TODO Underdiscussed: the theoretical result.

%maybe multiple sentences in def of It?
%explicit arcs making clear going back 1 word in the past, 2 words in the past? maybe matching between arcs and colors instead of the arrows?

%define math, AND give people intuition.

%
%\begin{figure}
%\includegraphics[width=\textwidth]{figures/theorem.png}
%	\caption{Left: }\label{fig:theorem}
%\end{figure}
%

%\caption{A speaker (left) utters a sentence. A comprehender (right) maintains a noisy representation of prior words, and incurs surprisal upon hearing the next word.}

Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages, using standard estimation methods for language modeling based on neural networks~\citep{hochreiter-long-1997}.
We compared these against chance, that is, to tradeoffs that would be expected for languages that apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
For each language, we constructed counterfactual versions that differ from real languages in their word order, holding all other properties constant.
These counterfactual versions use randomly created grammar for linearizing hierarchical structures into sentences (see Figure~\ref{fig:grammars}), akin to those found in human languages.
%These random grammars apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.

@@

- Potentially cite \cite{PhysRevLett.115.098701} as a reference saying that physical systems have bounded memory?

- Computational efficiency has been mentioned in mainstream syntax, though it's not so clear what it means there \citep{chomsky2005three,hauser2002faculty}.

@@

-This listener will represent information about past words in their short-term memory.
We quantify the amount of memory resources that are used by the number of bits.
We consider the question of how many bits of information this listener has to carry from one instant to the next.  %\jd{why?}

-Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
Understanding what a word means often requires consideration of context. 
Consider the sentence `The climber gave the attempt to reach the peak up.'
Understanding the word \emph{up} requires establishing that it is related to the word \emph{gave} -- neither of them has the meaning of `give up' by itself.
In order to understand this sentence correctly, a listener has to store some information about the occurrence of the word \emph{gave} until reaching \emph{up}.
For language comprehension, this means that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load. 
We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.

@@

% from the introduction to paper.tex:
%Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
%Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
%Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
%Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \citep{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
%
%At the same time, the field of functional linguistics has argued that these resource constraints not only affect online language processing, they also shape the form of human language itself.
%For example, the Performace--Grammar Correspondence Hypothesis (PGCH) of \citet{hawkins1994performance} holds that forms which are practically easier to produce and comprehend end up becoming part of the grammars of languages, and that this process can explain several of the universal properties of human languages originally documented by \citet{greenberg-universals-1963}.
%
%Here we take up the question of how to characterize short-term memory capacity limitations in language processing for both speakers and listeners, and the question of whether natural language grammars are shaped by these limitations.
%Whereas previous theories were based on specific mechanistic models of memory, our theory is purely information-theoretic, meaning that our predictions will hold independently across a wide variety of implementations and architectures.
%
%Our main new concept is the idea of a \emph{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language. Analogous results also hold for language production by resource-constrained speakers. We show evidence that the preferred word orders of natural languages are those that enable efficient memory--surprisal tradeoffs.
%
%

%However, testing this hypothesis has been difficult for two reasons:
%First, this requires a large amount of data representing the distribution of underlying hierarchical structures that people communicate,
%and
%second, because this presupposes a computatioal model of processing efficiency.
%Large-scale data has recently become available in the form of corpora annotated with underlying hierarchical structures across dozens of languages~\citep{nivre-universal-2017}.

@@

%
%Memory refers to information about previous parts of a sentence maintained in short-term memory.
%While memory limitations have been shown experimentally to affect human language processing, there is no consensus on the architecture used to represent memories.
%
%Here, for the first time, we provide an information-theoretic integration of these two aspects of processing, leading to a general theory of the computational efficiency of language.
%
%We denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the context $X_{<t}$.
%This listener will experience surprisal
%\begin{equation}
%	S :=	-\log P(X_t|L_t)
%\end{equation}
%upon hearing the next word.
%%We quantify memory as the number of bits about prior input that a listener represents.
%The minimal amount of information stored is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
%%related to work in physics on the ability of physical systems to represent memory, representing more bits costs more memory \citep{PhysRevLett.115.098701}
%We note that there is a tradeoff between average memory, $\E[S]$ and memory $H[L_t]$: A listener who invests more memory resources to create representations of the past will incur lower average surprisal.
%
%We predict that languages order information in such a way as to create an optimal tradeoff.

@@

\bibliographystyle{apalike}

@@

>>>>

\documentclass[12pt]{article}

\usepackage{showlabels}
\usepackage{fullpage}
\usepackage{pslatex}
%\usepackage{latexsym}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{url}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage{rotating}
\usepackage{amssymb}

@@

\frenchspacing

@@

\usepackage{scicite}

\usepackage{times}

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

\newenvironment{sciabstract}{%
\begin{quote} \bf}
{\end{quote}}

\date{}

%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%

@@

\baselineskip24pt

@@

%Since short-term memory of this kind is known to be highly limited in capacity \cite{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.

%The manuscript should start with a brief introduction describing the paper’s significance.
%The introduction should provide sufficient background information to make the article intelligible to readers in other disciplines, and sufficient context that the significance of the experimental findings is clear.
%Technical terms should be defined. Symbols, abbreviations, and acronyms should be defined the first time they are used.
%All tables and figures should be cited in numerical order. All data must be shown either in the main text or in the Supplementary Materials or must be available in an established database with accession details provided in the acknowledgements section. References to unpublished materials are not allowed to substantiate significant conclusions of the paper.

% introductory paragraph

@@

%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance};
however, it has been hard to test this idea without ad-hoc modeling assumptions.
%Alternative?: there has been debate about what shapes human language. role played by UG/innate constraints? environment? cognition?

We directly test this hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\cite{nivre-universal-2017}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}
\includegraphics[width=0.7\textwidth]{figures-gdrive/communication.png}
	\caption{Memory and Language Comprehension. (1) A speaker produces a sequence of words. (2) A listener maintains a representation of the words received so far. The listener can represent these at higher (left) or lower (right) levels of precision. (3) Throughout communication, the listener generates probabilistic expectations about the next word. Higher precision of memory representations leads to more precise predictions. (4) When receiving the next word, the listener incurs surprisal depending on the predictions. Higher levels of fidelity in memory lead to lower surprisal on average.
}\label{fig:communication}
\end{figure}

We consider the process of language comprehension in a person listening to an interlocutor, processing a stream of words uttered by the other person (Figure~\ref{fig:communication}).
Experimental research has established that such a listener (1) maintains information about the words received so far, and (2) forms probabilistic expectations about the upcoming words. %, and (3) words are easier to process when they are highly expected in context.
Importantly, the listener can represent the prior context at higher or lower levels of precision (Figure~\ref{fig:communication}).
More detailed memory representations lead to a higher working memory load, but also enable more precise predictions.
The psycholinguistic literature has established that humans differ in their capacity to store such representations, and that capacity limitations can create difficulties in language comprehension.

Experimental research has shown that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context.
Surprisal is known to be a strong linear predictor of processing effort as measured by per-word reading times.
For an idealized listener with unbounded memory resources, surprisal will be given by the negative log-probability of a word given its preceding context:
\begin{equation}\label{eq:surp}
	-\log p(x_i|x_{1...i-1})
\end{equation}
where $x_i$ is the $i$-th word in a given utterance (or discourse).
%
%In reality, a listener will have bounded capacity to represent prior context.
A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
As in Figure~\ref{fig:communication}, we denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
This listener will experience surprisal
\begin{equation}\label{eq:surp-listener}
	S :=	-\log p(x_t|L_t)
\end{equation}
upon hearing the next word $x_t$.
%The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
%For a listener who invests less bits into representing information $L_t$ about prior context, $L_t$ will be a more coarse-grained representation of past observations.
%Thus, the predictions made by such a listener will be on average less accurate, and such a listener will incur greater processing effort $S$.

Information theory, specifically the Data Processing Inequality, implies that the average surprisal $S$ will be higher for a listener who represents past input at lower levels of precision.
Consequently, there is a tradeoff between memory and surprisal:
A listener can achieve lower levels of processing effort, at the expense of increased memory load.

%Formally, we quantify memory as the number of bits about prior input that a listener represents.

@@

\begin{figure}
	(a)
\includegraphics[width=0.4\textwidth]{figures-gdrive/mi-distance.png}
	(b)
\includegraphics[width=0.2\textwidth]{figures-gdrive/theorem.png}
	(c)
\includegraphics[width=0.2\textwidth]{figures-gdrive/tradeoff.png}
	\caption{
		(a) Conditional mutual information $I_t$ describes how much predictive information about the next word is provided, on average, by the word $t$ steps in the past.
		(b) Theoretical result: We plot It (top) and tIt (bottom) as functions of t. A listener using B bits memory (bottom) to represent prior observations will incur at least A bits of extra surprisal beyond the entropy rate (top).
		(c) 
}\label{fig:theorem}
\end{figure}

%TODO
Let $I_t$ be the Conditional Mutual Information of two words $X_0, X_t$ conditioned on the intervening words:
\begin{equation}
	\operatorname{I}_t := \operatorname{I}[X_t, X_0 | X_1, \dots, X_{t-1}] = \operatorname{H}[X_t|X_1, \dots, X_{t-1}] - \operatorname{H}[X_t|X_0, \dots, X_{t-1}] 
\end{equation}
This quantity, visualized in Figure~\ref{fig:theorem} (a), measures how much predictive information is provided by the next word $X_t$ by the word $t$ steps in the past.
It is a statistical property of the language, and can be estimated from large-scale text data.
Our theoretical results describes how the tradeoff between memory and comprehension difficulty relates to $I_t$ (Figure~\ref{fig:theorem} (b)):
Consider a listener who invests $B$ bits of memory into representing past input.
We then consider the smallest $T$ such the area under the curve of $t I_t$, to the left of $T$, has size $B$.
Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
By tracing out all values $T >0$, one can obtain a bound on the tradeoff curve for any possible listener.

@@

%We prove that, for any integer $T > 0$, a listener encoding $\sum_{t=1}^T t I_t$ bits of memory will incur average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
%maybe multiple sentences in def of It?
%define math, AND give people intuition.

\begin{figure}
\includegraphics[width=0.7\textwidth]{figures-gdrive/counterfactual-languages.png}
	\caption{Estimating chance by constructing counterfactual grammars and languages.}\label{fig:grammars}
\end{figure}

Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages, using standard estimation methods for language modeling based on neural networks~\cite{hochreiter-long-1997}.
We compared these against chance, that is, to tradeoffs that would be expected for languages that apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
For each language, we constructed counterfactual versions that differ from real languages in their word order, holding all other properties constant.
These counterfactual versions use randomly created grammars for linearizing hierarchical structures into sentences (see Figure~\ref{fig:grammars}), akin to those found in human languages.
%These random grammars apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.

@@

%- Potentially cite \cite{PhysRevLett.115.098701} as a reference saying that physical systems have bounded memory?

- Computational efficiency has been mentioned in mainstream syntax, though it's not so clear what it means there \cite{chomsky2005three,hauser2002faculty}.

@@

%
%-This listener will represent information about past words in their short-term memory.
%We quantify the amount of memory resources that are used by the number of bits.
%We consider the question of how many bits of information this listener has to carry from one instant to the next.  %\jd{why?}
%
%-Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
%Understanding what a word means often requires consideration of context. 
%Consider the sentence `The climber gave the attempt to reach the peak up.'
%Understanding the word \emph{up} requires establishing that it is related to the word \emph{gave} -- neither of them has the meaning of `give up' by itself.
%In order to understand this sentence correctly, a listener has to store some information about the occurrence of the word \emph{gave} until reaching \emph{up}.
%For language comprehension, this means that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
%A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load. 
%We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.
%

@@

% from the introduction to paper.tex:
%Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
%Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
%Since short-term memory of this kind is known to be highly limited in capacity \cite{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
%Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \cite{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
%
%At the same time, the field of functional linguistics has argued that these resource constraints not only affect online language processing, they also shape the form of human language itself.
%For example, the Performace--Grammar Correspondence Hypothesis (PGCH) of \citet{hawkins1994performance} holds that forms which are practically easier to produce and comprehend end up becoming part of the grammars of languages, and that this process can explain several of the universal properties of human languages originally documented by \citet{greenberg-universals-1963}.
%
%Here we take up the question of how to characterize short-term memory capacity limitations in language processing for both speakers and listeners, and the question of whether natural language grammars are shaped by these limitations.
%Whereas previous theories were based on specific mechanistic models of memory, our theory is purely information-theoretic, meaning that our predictions will hold independently across a wide variety of implementations and architectures.
%
%Our main new concept is the idea of a \emph{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language. Analogous results also hold for language production by resource-constrained speakers. We show evidence that the preferred word orders of natural languages are those that enable efficient memory--surprisal tradeoffs.
%
%

%However, testing this hypothesis has been difficult for two reasons:
%First, this requires a large amount of data representing the distribution of underlying hierarchical structures that people communicate,
%and
%second, because this presupposes a computatioal model of processing efficiency.
%Large-scale data has recently become available in the form of corpora annotated with underlying hierarchical structures across dozens of languages~\cite{nivre-universal-2017}.

@@

%
%Memory refers to information about previous parts of a sentence maintained in short-term memory.
%While memory limitations have been shown experimentally to affect human language processing, there is no consensus on the architecture used to represent memories.
%
%Here, for the first time, we provide an information-theoretic integration of these two aspects of processing, leading to a general theory of the computational efficiency of language.
%
%We denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the context $X_{<t}$.
%This listener will experience surprisal
%\begin{equation}
%	S :=	-\log P(X_t|L_t)
%\end{equation}
%upon hearing the next word.
%%We quantify memory as the number of bits about prior input that a listener represents.
%The minimal amount of information stored is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
%%related to work in physics on the ability of physical systems to represent memory, representing more bits costs more memory \cite{PhysRevLett.115.098701}
%We note that there is a tradeoff between average memory, $\E[S]$ and memory $H[L_t]$: A listener who invests more memory resources to create representations of the past will incur lower average surprisal.
%
%We predict that languages order information in such a way as to create an optimal tradeoff.

@@

\bibliographystyle{Science}

@@

\appendix{SI}

\section{Materials and Methods}

\subsection{Experimental Design}
\subsection{Statistical Analysis}
